---
title: "How to Create an R Package"
author: "Stephen Knapp, Knapp Consulting LLC"
output: 
  html_document:
    toc: true
    toc_float: 
      toc_collapsed: true
    toc_depth: 4
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Introduction

This article provides an easy to follow checklist for creating an R package using the {devtools}/{usethis} framework as explained in the book [R Packages (2e)](https://r-pkgs.org/) by Hadley Wickham and Jennifer Bryan. The target audience for this article is for both new and experienced R developers. For new R developers, creating an R package may seem overwhelming and complicated at first but, this article aims to remove the mystery from the process by providing a tutorial. Experienced R developers who need a quick refresher on how to set-up an R package or add certain features to an existing R package (such as a {pkgdown} website) can use this article as a quick reference guide or cheat sheet.

This article will walk you through how to build a professional R package with the following features:

 - **Version control using git and GitHub:**
 
   - Automated GitHub actions for R CMD check with README badge
   
   - Automated documentation and code styling with pull requests
 
 - **Testing:**
 
   - Unit testing using {testthat}
 
   - Automated GitHub actions for code testing coverage using CodeCov and {covr} with README badge
   
 - **Professional looking package website hosted on GitHub Pages and built using {pkgdown}:**
 
   - Includes instructions on how make a package logo
 
 - **Internal and external data files**
 
 - **Documentation:**
   
   - Function and data documentation using {roxygen2}
   
   - README using a .Rmd file
   
   - NEWS.md
   
   - Articles and vignettes
   
   - DESCRIPTION file set-up with license
 
 - **Code styling using {styler}**
 
The checklist consists of 32 steps in total which are divided in to 10 sections, A through J. Where possible, I provide a rough time estimate to complete each section: 

 - The first four sections A. Initial Set-up, B. Version Control Using git and GitHub, C. Package Documentation, and D. GitHub Actions, which collectively encompass steps 1 through 11, are considered the minimum required to establish a proper R package framework and take roughly 30 minutes to complete. 
 
 - Section E. Functions and Function Documentation includes steps 12, 13, and 14 and involves writing the actual R code for your package; how long this takes depends on the package.
 
 - Note: once you have completed sections A through D (steps 1 through 11) and as your work on section E (steps 12 through 14) you should start using the standard development work flow described in the Post Set-up Workflow section of this article.
 
 - All sections after E. Functions and Function Documentation are optional and can be completed as needed or as time allows, in any order. These sections include:
 
   + F. Code Styling (step 15). Time commitment: about 5 minutes.
   
   + G. Additional Package Documentation (step 16). Time commitment: depends on how many package vignettes and articles you want.
   
   + H. Data Files (steps 17 through 21). Time commitment: depends on how many data files you need.
   
   + I. Testing (steps 22 through 28). Time commitment: depends on the amount of code in your package and what level of code test coverage you consider acceptable.
   
   + J. Website and Logo (steps 29 through 32). Time commitment: about 30 minutes.
   
Each step in this article also provides references with links.

 
### Prerequisites

**GitHub and CodeCov accounts:** throughout the article certain platforms are chosen to perform certain actions: GitHub for repository hosting, GitHub Actions for CI/CD pipelines, GitHub Pages for the {pkgdown} website, and CodeCov for unit testing coverage. There are alternatives to these platforms however, the {devtools}, {usethis}, {pkgdown}, {covr} packages used to build these features are conducive to using these popular platforms therefore, you must have GitHub and CodeCov accounts.

**R Studio Installation Settings:** this article assumes your R Studio installation was previously configured according to [the {usethis} set-up guide](https://usethis.r-lib.org/articles/usethis-setup.html). This guide ensures that {devtools} and {usethis} are loaded every time you start R Studio, it provides default DESCRIPTION file settings, and connects your GitHub account to R Studio.
 

## Steps

### A. Initial Set-up

#### 1. Check the package name is available

```{r}
available::available("[PROPOSED PACKAGE NAME]")
```

Reference: [https://r-pkgs.org/workflow101.html#use-the-available-package](https://r-pkgs.org/workflow101.html#use-the-available-package)

#### 2. Create a new directory

```{r}
dir.create("[PATH TO PACKAGE]")
```


#### 3. Set-up the directory as an R package:

```{r}
setwd("[PATH TO PACKAGE]")
create_package()
```

Reference: [https://r-pkgs.org/whole-game.html#create_package](https://r-pkgs.org/whole-game.html#create_package)


### B. Version Control using git and GitHub

#### 4. Set-up local version control

```{r}
use_git()
```

Next, make your first commit using the terminal using the git commands add and commit.

Reference: [https://r-pkgs.org/whole-game.html#use_git](https://r-pkgs.org/whole-game.html#use_git)

#### 5. Create a GitHub remote repository

```{r}
# if you did step 5, ignore the first message about uncommitted changes
use_github()
```

Reference: [https://r-pkgs.org/whole-game.html#use_github](https://r-pkgs.org/whole-game.html#use_github)


### C. Package Documentation

#### 6. Populate the DESCRIPTION file

Update these fields:
 
 - Title
 
 - Authors (this should be auto-populated if you followed the [the {usethis} set-up guide](https://usethis.r-lib.org/articles/usethis-setup.html))
 
 - Description
 
 - Chose a license:

```{r}
# the MIT license is one of the most permissive but there are other options
use_mit_license()
```

References: 

 - DESCRIPTION file: [https://r-pkgs.org/description.html](https://r-pkgs.org/description.html)
 
 - Licensing: [https://r-pkgs.org/license.html](https://r-pkgs.org/license.html)

#### 7. Create a README using Rmarkdown

```{r}
use_readme_rmd()
```

After populating the basic information in the README.Rmd file, build the .md version of the file using:

```{r}
build_readme()
```

Reference: [https://r-pkgs.org/whole-game.html#use_readme_rmd](https://r-pkgs.org/whole-game.html#use_readme_rmd)

#### 8. Create a NEWS file

```{r}
use_news_md()
```

Edit the file as needed. It will be used as a change log for package version releases.

Reference: [https://r-pkgs.org/other-markdown.html#sec-news](https://r-pkgs.org/other-markdown.html#sec-news)

#### 9. Create package documentation

This creates a dummy file to establish package level documentation:

```{r}
use_package_doc()
```

Reference: [https://r-pkgs.org/man.html#sec-man-package-doc](https://r-pkgs.org/man.html#sec-man-package-doc)
 

### D. GitHub Actions

#### 10. Automate R CMD check

Every time you push the package to the remote repository, R CMD check will be run and system compatibility will be checked for a variety of OS's. 

```{r}
use_github_action("check-standard")
```

This will also add the R CMD check badge to the README.Rmd file so the README.md file should be re-built:

```{r}
build_readme()
```

Reference: [https://r-pkgs.org/software-development-practices.html#r-cmd-check-via-gha](https://r-pkgs.org/software-development-practices.html#r-cmd-check-via-gha)


#### 11. Automate documentation and code styling for pull requests

```{r}
use_github_action("pr-commands")
```

Reference: [https://usethis.r-lib.org/reference/github_actions.html#use-github-action-pr-commands-](https://usethis.r-lib.org/reference/github_actions.html#use-github-action-pr-commands-)

### E. Functions and Function Documentation

#### 12. Add your code

Your package code must exist in .R files in the `R/` directory. Create each .R file using:

```{r}
use_r("[NEW R SCRIPT NAME]")
```

Reference: [https://r-pkgs.org/whole-game.html#use_r](https://r-pkgs.org/whole-game.html#use_r)

#### 13. Import packages

As functions from other packages are used throughout your code they must always be called using the `[PACKAGE NAME]::[FUNCTION]()` method. Each unique package that is called in your code must be added as an import in the DESCRIPTION file. This can be done using:

```{r}
use_package("[IMPORTED PACKAGE NAME]")
```

It's also a good idea to import {magritts}'s pipe operator:

```{r}
use_pipe(export = TRUE)
```


References: 

 - [https://r-pkgs.org/whole-game.html#use_package](https://r-pkgs.org/whole-game.html#use_package)
 
 - [https://usethis.r-lib.org/reference/use_pipe.html](https://usethis.r-lib.org/reference/use_pipe.html)


#### 14. Create function documentation

For each function in the R/ directory, create {roxygen2} comments using the references below, then build all the package documentation files by running:

```{r}
document()
```

References:

 - [https://r-pkgs.org/man.html#sec-man-package-doc](https://r-pkgs.org/man.html#sec-man-package-doc)
 
 - [https://r-pkgs.org/man.html](https://r-pkgs.org/man.html)
 
 - [{roxygen2}](https://roxygen2.r-lib.org/)
 
### F. Code Styling
 
#### 15. Apply code styling using the {styler} package

To style the entire package run:

```{r}
use_tidy_style()
```

Note: you should always commit changes immediately before and after running the above code to separate style changes from code changes in the git history.

References:

 - [https://r-pkgs.org/code.html#code-style](https://r-pkgs.org/code.html#code-style)
 
 - [{styler}](https://styler.r-lib.org/)
 
### G. Additional Package Documentation

#### 16. Articles and vignettes

Use articles and vignettes to provide package instructions and example workflows beyond what is in README.Rmd. Vignettes will be included with your package on install. Articles will not be included with your package on installation but will be available on the {pkgdown} website. These can be created using the functions:

```{r}
use_vignette("[NEW VIGNETTE NAME]")
use_article("[NEW ARTICLE NAME]")
```

The first time each of these functions is called the directories `vignettes` and `vignettes/articles` are created, respectively.
 
Reference: [https://r-pkgs.org/vignettes.html](https://r-pkgs.org/vignettes.html)

### H. Data Files

There are five locations for storing/creating R package data files. Choosing a directory to use for a file depends on how the file is to be used and whether the user should have access to the file after they install your package.

1. `data-raw/`: this directory can store raw data files and is used for creating data in any of the other data directories below. Nothing in `data-raw/` will be available to end users after installation and `load_all()` does not load any data files in this directory into the developer's environment. See step 17.

2. `data/`: this directory stores data files, in a native R format (ie .rda), that the user will have access to after package installation after calling [PACKAGE NAME]::[DATA OBJECT NAME] or by calling `library([PACKAGE NAME])` and then referring to the DATA OBJECT NAME. Each data object should have its own .rda file and must be documented. See steps 18 and 19. As the package developer, can load these data files into your environment using `load_all()`. 

3. `R/sysdata.rda`: this is a file that stores all R data objects for internal package use. These data objects will not be available to the user after installation. See step 20. As the package developer, can load these data files into your environment using `load_all()`. 

4. `inst/extdata`: this directory stores data files that should be available to the user after installation but which are not in a native R format (ie .csv). See step 21. These files cannot be called directly as R objects by the end user and `load_all()` will not make these available in the developer's environment. The package developer or end user can get the path to a data file in this directory by calling `system.file("extdata/[DATA FILE NAME]", package = "[PACKAGE NAME]")` and then load that file into the global environment by calling the appropriate read function (ie `read.csv()`).

5. `tests/testthat/fixtures`: this directory stores data files in any format that are used for unit testing. See the 'Testing' section, specifically step 25, of this article for more information. By default, these files can be accessed by the user after installation. These files cannot be called directly as R objects by the end user and `load_all()` will not make these available in the developer's environment. The package developer can get the path to a data file in this directory by calling `testthat::test_path("[DATA FILE NAME]")` and then load that file into the global environment by calling the appropriate read function (ie `read.csv()`). An end users would need to specify the full path to the file in their R library directory and then call the appropriate read function.


#### 17. Set-up the raw data directory

The function below will create an R script for creating data files. The first time it is called, it will also create the `data-raw/` directory.

```{r}
use_data_raw("[NEW DATA FILE NAME]")
```

It's good practice to have one R script in `data-raw/` for each data file in the package. At the end of each script, you will save the data object to one of the other four data locations.

Reference: [https://r-pkgs.org/data.html#sec-data-data-raw](https://r-pkgs.org/data.html#sec-data-data-raw)


#### 18. Create an exported data file in data\

This step will create a file in the `data\` directory which will be available to a user after package installation. Data stored in this manner must be in a native R data format (ie `.rda`). For each script in `data-raw/` that is supposed to create an exported data file, call the function below at the end to save the file in the correct location:

```{r}
use_data("[NEW DATA FILE NAME]", overwrite = TRUE)
```

Reference: [https://r-pkgs.org/data.html#sec-data-data](https://r-pkgs.org/data.html#sec-data-data)

#### 19. Document a data file in the data/ directory

Every file in `data/` must have package documentation created using {roxygen2}. The documentation for all of the files in `data/` should reside in the file `R/data.R` file. Create the file using:

```{r}
use_r("data.R")
```

Then populate it per the reference below.

Reference: [https://r-pkgs.org/data.html#sec-documenting-data](https://r-pkgs.org/data.html#sec-documenting-data)


#### 20. Create package internal data

Internal package data will not be shared with the end users of the package. All internal data is stored in the R object `R/sysdata.rda`. You can create a script to create this object in `data-raw` like this:

```{r}
use_data_raw("sysdata")
```

At the end of the script call:

```{r}
use_data(..., overwrite = TRUE, internal = TRUE)
```

Reference: [https://r-pkgs.org/data.html#sec-data-sysdata](https://r-pkgs.org/data.html#sec-data-sysdata)

#### 21. Use non-R data files

Any non-R data files that you want the user to have access to after installation should be stored in `inst/extdata`. They can be created under `data-raw` but at the end of the script should not call `use_data()`. Instead save them using something like:

```{r}
dir <- system.file("extdata", package = "[PACKAGE NAME]")
write.csv(data, file = file.path(dir, "[FILE NAME].csv"))
```

Reference: [https://r-pkgs.org/data.html#sec-data-extdata](https://r-pkgs.org/data.html#sec-data-extdata)


### I. Testing

#### 22. Set-up {testthat}

```{r}
use_testthat()
```

Reference: [https://r-pkgs.org/testing-basics.html#initial-setup](https://r-pkgs.org/testing-basics.html#initial-setup)

#### 23. Writing tests

Create a test script for each file in `R/` that you want to test using:

```{r}
use_test("[NAME OF .R FILE IN R/]")
```

Populate the file according to the {testthat} convention using the references below.

References:

 - [https://r-pkgs.org/testing-basics.html#sec-tests-mechanics-workflow](https://r-pkgs.org/testing-basics.html#sec-tests-mechanics-workflow)
 
 - [{testthat}](https://testthat.r-lib.org/)


#### 24. Running tests

You can run individual test functions inside a test file by highlighting and running that section of code. Make sure to always run `load_all()` first and after making any change to the file being tested.

You can run all tests for a file in `R` using:

```{r}
test_file("[NAME OF .R FILE IN R/]")
```

Test the entire package by running:

```{r}
test()
```


References:

 - [https://r-pkgs.org/testing-basics.html#sec-tests-mechanics-workflow](https://r-pkgs.org/testing-basics.html#sec-tests-mechanics-workflow)
 
 - [{testthat}](https://testthat.r-lib.org/)


#### 25. Testing helper functions and fixtures

You can include test helper functions by creating the file `tests/testthat/helpers.R`.  

You can also include data files to assist with testing (called test fixtures) creating the directory `tests/testthat/fixtures` and storing them there. These can be native R files or not. It's a good idea to create these data files in the `data-raw/` directory (see section H. Data Files).

References:

 - [https://r-pkgs.org/testing-basics.html#sec-tests-mechanics-workflow](https://r-pkgs.org/testing-basics.html#sec-tests-mechanics-workflow)
 
 - [https://r-pkgs.org/testing-advanced.html#test-fixtures](https://r-pkgs.org/testing-advanced.html#test-fixtures)
 

#### 26. Prepare the testing coverage framework for CodeCov

```{r}
use_coverage(type = "codecov")
```

The code above will also add a CodeCov badge the README so the README must be re-built:

```{r}
build_readme()
```

Reference: [https://usethis.r-lib.org/reference/use_coverage.html](https://usethis.r-lib.org/reference/use_coverage.html)


#### 27. Check the test coverage

To check test coverage for the entire package run one of the two lines of code below, depending the output type you want to see. Note before running the functions below, you must restart R.

```{r}
# this will produce a report in R Studio's Viewer window
covr::report()

# alternatively, this will print the results to the console
covr::package_coverage()
```

To check test coverage for an individual file run the following line of code which will produce a report in the R Studio Viewer window that will highlight which lines of code are tested and which are not tested.

```{r}
devtools::test_coverage_active_file()
```

References:
 
 - [{covr}](https://covr.r-lib.org/) for unit testing automation
 
 - [https://r-pkgs.org/testing-design.html#sec-testing-design-coverage](https://r-pkgs.org/testing-design.html#sec-testing-design-coverage)

#### 28. Link the GitHub repo to CodeCov

Sign into CodeCov, select the correct repo, and following the steps for GitHub Actions set-up. This will involve copying a token from CodeCov into a GitHub repo secret. Note the GitHub repo must be public for this to work. 

Finally, add test coverage as a GitHub actions workflow:

```{r}
use_github_action("test-coverage")
```

References:

 - [https://usethis.r-lib.org/reference/use_coverage.html](https://usethis.r-lib.org/reference/use_coverage.html)
 
 - [{covr}](https://covr.r-lib.org/) for unit testing automation
 
 - [https://r-pkgs.org/testing-design.html#sec-testing-design-coverage](https://r-pkgs.org/testing-design.html#sec-testing-design-coverage)


### J. Website and Logo

#### 29. Set-up a {pkgdown} website using GitHub Pages

GitHub pages will be used to host the website. Every time you push to the remote, the website will be re-built.

```{r}
use_pkgdown_github_pages()
```

References:

 - [https://r-pkgs.org/website.html#sec-website-deployment](https://r-pkgs.org/website.html#sec-website-deployment)
 
 - [{pkgdown}](https://pkgdown.r-lib.org/)

#### 30. Build and preview the site locally

To initially create the files for building the site locally run:

```{r}
pkgdown::build_site()
```

To locally preview sections of the site run one of the `pkgdown::build_**()` functions.

References:

 - [https://r-pkgs.org/website.html#sec-website-deployment](https://r-pkgs.org/website.html#sec-website-deployment)
 
 - [{pkgdown}](https://pkgdown.r-lib.org/)


#### 31. Customize the reference index (optional)

By default the reference page of the site will list all package functions in alphabetical order. To change this page and group packages into sections, see the reference below.

Reference: [https://pkgdown.r-lib.org/reference/build_reference.html?q=subtitle#reference-index](https://pkgdown.r-lib.org/reference/build_reference.html?q=subtitle#reference-index)

#### 32. Create a hex logo (optional)

The package logo will be displayed in your README file and on the {pkgdown} website and really makes your package stand-out and look professional. 

Design your logo using [hexmake](https://connect.thinkr.fr/hexmake/) and download the .png file to a directory outside of your package. Move your logo into your package and configure it using:

```{r}
use_logo("[PATH TO LOGO]")
```

The code ran in the previous chunk will add html code to your clipboard. Paste that into README.Rmd where the title is. Then rebuild the README:

```{r}
build_readme()
```

Reference: [https://r-pkgs.org/website.html#logo](https://r-pkgs.org/website.html#logo)


## Post Set-up workflow

The iterative workflow steps are generally to develop your package code, re-load the package into the global environment, test it, document it, run R CMD check, then commit and push your changes to GitHub. The specific steps are outlined below.

Iteratively develop the code in one file in `R/` using:

```{r}
# update global environment with recent package changes
load_all()

# test the file being worked on by running sections of the test file or by running:
test_active_file()
```

Once you are satisfied with the changes, test the entire package using:

```{r}
test()
```

Update the package documentation using:

```{r}
document()

# when changes to the README are made, also run
build_readme()
```

Run R CMD check:

```{r}
check()
```

If R CMD check passes, with 0 errors and 0 warnings, commit the changes to the repo using the standard git commands add, commit, and push in the terminal.

Reference: [{devtools} cheatsheet](https://devtools.r-lib.org/)


## References

The primary reference used is the book [R Packages (2e)](https://r-pkgs.org/) by Hadley Wickham and Jennifer Bryan. Other references include:

 - [{usethis}](https://usethis.r-lib.org/index.html) for package development tools
 
 - [{devtools}](https://devtools.r-lib.org/) and the {devtools} cheatsheet for package development tools
 
 - [{testthat}](https://testthat.r-lib.org/) for unit tests
 
 - [{covr}](https://covr.r-lib.org/) for unit testing automation
 
 - [{roxygen2}](https://roxygen2.r-lib.org/) for documentation
 
 - [{pkgdown}](https://pkgdown.r-lib.org/) for building package websites
 
 - [{styler}](https://styler.r-lib.org/) for code styling

















